---
title: "exercise-10"
author: "John Hinkle"
format: html
editor: visual
self-contained: true
---

## Preliminaries

```{r}
library(tidyverse) # tidyverse suite of packages
library(ggplot2) # for plotting
library(cowplot) # for making multi-panel figures
library(skimr) # for exploratory data analysis

d <- read_csv("https://raw.githubusercontent.com/difiore/ada-datasets/refs/heads/main/AVONETdataset1.csv")

# winnow dataset to the specified variables
d <- d |> select(Species1, Family1, Order1, Beak.Length_Culmen, Beak.Width, Beak.Depth, Tarsus.Length, Wing.Length, Tail.Length, Mass, Habitat, Migration, Trophic.Level, Trophic.Niche, Min.Latitude, Max.Latitude, Centroid.Latitude, Primary.Lifestyle, Range.Size)

skim(d)
```

# Challenge 1

## Step 1: Boxplots of log(**Mass**) in relation to **trophic level and migration behavior type**

```{r}
d <- d |> 
  mutate(Migration = as.factor(Migration)) |> 
  mutate(logMass = log(Mass)) |>
  mutate(Trophic.Level = as.factor(Trophic.Level))

p1 <- ggplot(data = d |> drop_na(Trophic.Level), aes(x = Trophic.Level, y = logMass)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.1, color = "dodgerblue4")

p2 <- ggplot(data = d |> drop_na(Migration), 
       aes(x = Migration, y = logMass)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.1, color = "dodgerblue4")

plot_grid(p1, p2, nrow = 2, labels = c("I)", "II)"))
```

## Step 2a: Linear models of log(**Mass**) in relation to **trophic level, and in relation to migration behavior type**

Based on the model, body mass is not significantly different between carnivores and omnivores. Body mass is significantly different between the other trophic levels (i.e., carnivore vs. herbivore, carnivore vs. scavenger, omnivore vs. herbivore, omnivore vs. scavenger)

```{r}
# "Carnivore" as reference
m1a <- lm(logMass ~ Trophic.Level, data = d)
summary(m1a)
```

```{r}
# Relevel to make "Herbivore" the reference
d$Trophic.Level <- relevel(d$Trophic.Level, ref = "Herbivore")
m1b <- lm(logMass ~ Trophic.Level, data = d)
summary(m1b)
```

```{r}
# Relevel to make "Ominivore" the reference
d$Trophic.Level <- relevel(d$Trophic.Level, ref = "Omnivore")
m1c <- lm(logMass ~ Trophic.Level, data = d)
summary(m1c)
```

```{r}
# Relevel to make "Scavenger" the reference
d$Trophic.Level <- relevel(d$Trophic.Level, ref = "Scavenger")
m1d <- lm(logMass ~ Trophic.Level, data = d)
summary(m1d)
```

## Step 2b: Linear models of log(**Mass**) in relation to **migration behavior type**

Based on the model, body mass is significantly different between each migration behavior type.

```{r}
# "1" as reference
m2a <- lm(logMass ~ Migration, data = d)
summary(m2a)
```

```{r}
# Relevel to make "2" the reference
d$Migration <- relevel(d$Migration, ref = "2")
m2b <- lm(logMass ~ Migration, data = d)
summary(m2b)
```

```{r}
# Relevel to make "3" the reference
d$Migration <- relevel(d$Migration, ref = "3")
m2c <- lm(logMass ~ Migration, data = d)
summary(m2c)
```

## Step 3: Perform post-hoc Tukey Honest Significant Differences test for migration behavior types

Conducting a post-hoc Tukey HSD test reveals the same results seen in Step 2b. Body mass is significantly different between each migration behavior type.

```{r}
m1aov <- aov(logMass ~ Migration, data = d)
(posthoc <- TukeyHSD(m1aov, which = "Migration", conf.level = 0.95))
```

## Step 4: Permutation

```{r}
library(infer) # for permutation

original.F <- aov(logMass ~ Trophic.Level, data = d) |>
  broom::tidy() |>
  filter(term == "Trophic.Level")

permuted.F <- d |>
  specify(logMass ~ Trophic.Level) |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate(stat = "F")

visualize(permuted.F) + shade_p_value(obs_stat = original.F$statistic, direction = "greater")
```

```{r, warning=FALSE}
p.value <- permuted.F |>
    get_p_value(obs_stat = original.F$statistic, direction = "greater")
                
# original p-value
paste("The original p-value is:", original.F$p.value)

# p-value based on permutation
# p = 0, but I wonder if that is due to rounding as the original p-value is very small
paste("The p-value based on the permutation is:",p.value[["p_value"]])
```

# Challenge 2

## Step 1: Create relative beak length and relative tarsus length variables

```{r}
m3 <- lm(log(Beak.Length_Culmen) ~ logMass, data = d)
m4 <- lm(log(Tarsus.Length) ~ logMass, data = d)

d <- d |> mutate(RelativeBeakLength = m3$residuals) |> 
  mutate(RelativeTarsusLength = m4$residuals)
```

## Step 2: Box plots of I) Relative beak length variable in relation to trophic niche and II) Relative tarsus length variable in relation to primary lifestyle

```{r}
box1 <- ggplot(data = d, aes(x = Trophic.Niche, y = RelativeBeakLength)) +
  geom_boxplot(color = "dodgerblue4") +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

box2 <- ggplot(data = d, aes(x = Primary.Lifestyle, y = RelativeTarsusLength)) +
  geom_boxplot(color = "dodgerblue4") +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

plot_grid(box1, box2, nrow = 1, labels = c("I)", "II)"))
```

## Step 3a: Association between geographic range size and migration behavior type

```{r}
migration <- d |> 
  drop_na(Migration) |>
  mutate(logRS = log(Range.Size))

h1 <- ggplot(migration, aes(x = Range.Size)) +
  geom_histogram(fill= "#bf5700")

# log transform range size
h2 <- ggplot(migration, aes(x = logRS)) +
  geom_histogram(fill= "#bf5700")

plot_grid(h1, h2, nrow = 2, labels = c("I)", "II)"))
```

## Step 3b: Run models for log(Range.Size) vs. Migration

The p-values (p\<0.001) indicate that there is a significant difference in the log(Range.Size) of each migration behavior type. It seems reasonable that a migratory bird would have a larger range size than a non-migratory bird.

```{r}
#  "3" as the reference
m5a <- lm(logRS ~ Migration, data = migration)
summary(m5a)
```

```{r}
# Relevel to make "2" the reference
migration$Migration <- relevel(migration$Migration, ref = "2")
m5b <- lm(logRS ~ Migration, data = migration)
summary(m5b)
```

```{r}
# Relevel to make "1" the reference
migration$Migration <- relevel(migration$Migration, ref = "1")
m5c <- lm(logRS ~ Migration, data = migration)
summary(m5c)
```

### Post-hoc Tukey HSD test

The post-hoc Tukey HSD test also indicates that there is a significant difference in log(Range.Size) between each of the migration behavior types.

```{r}
m5aov <- aov(logRS ~ Migration, data = migration)
(posthoc <- TukeyHSD(m5aov, which = "Migration", conf.level = 0.95))
```

## Step 4a: Analyze Infraorder *Passeriformes* (aka songbirds)

```{r}
songbirds <- d |> filter(Order1 == "Passeriformes")

box1 <- ggplot(data = songbirds, aes(x = Primary.Lifestyle, y = RelativeBeakLength)) +
  geom_boxplot() +
  geom_jitter(color = "#bf5700", alpha = 0.1) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

box2 <- ggplot(data = songbirds, aes(x = Primary.Lifestyle, y = RelativeTarsusLength)) +
  geom_boxplot() +
  geom_jitter(color = "#bf5700", alpha = 0.1) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

plot_grid(box1, box2, nrow = 1, labels = c("I)", "II)"))
```

## Step 4b: Models

The p-values (p\<0.001) indicate that there is a significant difference in relative beak length and relative tarsus length between each primary lifestyle.

```{r}
# Linear model for relative beak length vs. primary lifestyle
m6 <- lm(RelativeBeakLength ~ Primary.Lifestyle, data = songbirds)
summary(m6)

# Linear model for relative tarsus length vs. primary lifestyle
m7 <- lm(RelativeTarsusLength ~ Primary.Lifestyle, data = songbirds)
summary(m7)
```

## Step 5: Two-way model

The p-values (p\<0.001) indicate that there is a significant difference in relative beak length between each primary lifestyle and trophic level.

```{r}
# Two-way model
m8 <- lm(RelativeBeakLength ~ Primary.Lifestyle + Trophic.Level, data = songbirds)
summary(m8)
```

## Step 6: Two-way model plus an interaction term (Primary.Lifestyle:Trophic.Level)

```{r}
m9 <- lm(RelativeBeakLength ~ Primary.Lifestyle + Trophic.Level + Primary.Lifestyle:Trophic.Level, data = songbirds)
summary(m9)

# Another way to do this would be: 
# m9 <- lm(RelativeBeakLength ~ Primary.Lifestyle + Primary.Lifestyle*Trophic.Level, data = songbirds)
```

## Step 7: Interaction plot

```{r}
interaction.plot(x.factor = songbirds$Primary.Lifestyle, xlab = "Primary.Lifestyle", trace.factor = songbirds$Trophic.Level, trace.label = "Trophic.Level",
    response = songbirds$RelativeBeakLength, fun = base::mean, ylab = "RelativeBeakLength")
```

## Step 8: Checking if the data meet the primary assumptions for standard linear regression and ANOVA

It looks like all of our data does not meet the primary assumptions with some relationships we looked at violating the "max/min ratio \< 2" rule of thumb.

```{r}

#logRS ~ Migration
stats1 <- d |>
    group_by(Trophic.Level) |>
    summarize(`mean(logMass)` = mean(logMass), `sd(logMass)` = sd(logMass))

stats2 <- d |>
    group_by(Migration) |>
    summarize(`mean(logMass)` = mean(logMass), `sd(logMass)` = sd(logMass))

stats3 <- d |>
    group_by(Trophic.Level) |>
    summarize(`mean(RelativeBeakLength)` = mean(RelativeBeakLength), `sd(RelativeBeakLength)` = sd(RelativeBeakLength))

stats4 <- d |>
    group_by(Primary.Lifestyle) |>
    summarize(`mean(RelativeTarsusLength)` = mean(RelativeTarsusLength), `sd(RelativeTarsusLength)` = sd(RelativeTarsusLength))

stats5 <- songbirds |>
    group_by(Trophic.Level) |>
    summarize(`mean(RelativeBeakLength)` = mean(RelativeBeakLength), `sd(RelativeBeakLength)` = sd(RelativeBeakLength))

# check that variances are roughly equal (ratio of max/min is <2)
paste("logMass ~ Trophic.Level:", max(stats1$`sd(logMass)`)/min(stats1$`sd(logMass)`))

paste("logMass ~ Migration:", max(stats2$`sd(logMass)`)/min(stats2$`sd(logMass)`))

paste("RelativeBeakLength ~ Trophic.Level:", max(stats3$`sd(RelativeBeakLength)`)/min(stats3$`sd(RelativeBeakLength)`))

paste("RelativeTarsusLength ~ Primary.Lifestyle:", max(stats4$`sd(RelativeTarsusLength)`)/min(stats4$`sd(RelativeTarsusLength)`))

paste("RelativeBeakLength ~ Trophic.Level (songbirds):", max(stats5$`sd(RelativeBeakLength)`)/min(stats5$`sd(RelativeBeakLength)`))
```
